\Large\bf FlexibleSUSY --- A spectrum generator generator for supersymmetric models

### 2014-11-06 doc/flexiblesusy-paper.tex
adding sdown, selectron, neutralino to pole mass table
align columns to the right
discuss Table 1 in more detail

The (physical) pole mass spectrum can now be calculated.  \fs uses the
full one-loop self-energies and tree-level mass matrices obtained from
SARAH to calculate the pole masses, which means finding the values $p$
that solve the equation
%
\begin{align}
  0 = \det\left[p^2\unity - m_{f,1L}(p^2)\right].
  \label{eq:pole-mass-def}
\end{align}
%
Here the one-loop mass matrix $m_{f,1L}(p^2)$ for field $f$ is given
in terms of the tree-level mass matrix $m_f$ and the self-energy
$\Sigma_f(p^2)$ as
%
\begin{align}
  &\text{scalars } \phi: &
  m_{\phi,1L}(p^2) &= m_{\phi} - \Sigma_\phi(p^2), \\
  &\text{Majorana fermions } \chi: &
  m_{\chi,1L}(p^2) &= m_{\chi} - \frac{1}{2}\Big[
    \Sigma_\chi^S(p^2) + \Sigma_\chi^{S,T}(p^2)
    + \Big( \Sigma_\chi^{L,T}(p^2) + \Sigma_\chi^R(p^2) \Big) m_{\chi} \notag \\
    &&&\phantom{= m_{\chi} - \frac{1}{2}\Big[}
    + m_{\chi} \Big( \Sigma_\chi^L(p^2) + \Sigma_\chi^{R,T}(p^2) \Big)
  \Big], \\
  &\text{Dirac fermions } \psi: &
  m_{\psi,1L}(p^2) &= m_{\psi}
  - \Sigma_\psi^S(p^2)
  - \Sigma_\psi^R(p^2) m_{\psi}
  - m_{\psi} \Sigma_\psi^L(p^2) .
\end{align}
%
Eq.~\eqref{eq:pole-mass-def} can be solved by diagonalizing the
one-loop mass matrix $m_{f,1L}(p^2)$.  However, since $m_{f,1L}(p^2)$
depends on the momentum $p$, an iteration over $p$ must be performed.
Since this iteration can be very time consuming for large field
multiplets, \fs provides two approximative procedures with a shorter
run-time in addition to the iterative procedure.  In the \fs model
file the two approximative procedures are called \code{LowPrecision}
and \code{MediumPrecision}.  The iterative procedure is called
\code{HighPrecision}.  The procedure to be used can be set in the
model file for each field.  The default setting is
%
\begin{lstlisting}
DefaultPoleMassPrecision = MediumPrecision;
HighPoleMassPrecision    = {hh, Ah, Hpm};
MediumPoleMassPrecision  = {};
LowPoleMassPrecision     = {};
\end{lstlisting}
%
In the variable \code{DefaultPoleMassPrecision} the default pole mass
calculation precision to be used for all particles is selected.
Possible values are \code{LowPrecision}, \code{MediumPrecision} and
\code{HighPrecision}.  The values \code{LowPrecision} and
\code{MediumPrecision} correspond to the two approximations described
below and \code{HighPrecision} corresponds to the iterative
determination of the pole masses.  In the variables
\code{HighPoleMassPrecision}, \code{MediumPoleMassPrecision} and
\code{LowPoleMassPrecision} the pole mass calculation precision can be
changed for individual particles.  The settings in these variables
overwrite the setting in \code{DefaultPoleMassPrecision} for these
particles.  In the above example the pole masses of all particles are
calculated with \code{MediumPrecision}, except for the Higgs boson
pole masses, which are calculated with the iterative procedure
(\code{HighPrecision}).

The three different pole mass calculation procedures work as follows:
%
\begin{itemize}
\item \code{LowPrecision}: This option provides the lowest precision
  but is also the fastest one.  Here the one-loop mass matrix
  $m_{f,1L}^\text{low}$ is calculated exactly once as
%
  \begin{align}
    \forall i,j: (m_{f,1L}^\text{low})_{ij} = (m_{f,1L}(p^2 = m_{f_i}
    m_{f_j}))_{ij} ,
  \end{align}
%
  where $m_{f_i}$ is the $i$th mass eigenvalue of the tree-level mass
  matrix $m_f$.  Afterwards, $m_{f,1L}^\text{low}$ is diagonalized and
  the eigenvalues are interpreted as pole masses $m_{f_i}^\pole$.
  This method neglects terms of the form
  %
  \begin{align}
    \left[(m_{f_k}^\pole)^2 - m_{f_i}m_{f_j}\right]
    \left.\frac{\partial m_{f,1L}(p^2)}{\partial
        p^2}\right|_{p^2=m_{f_i}m_{f_j}} ,
  \end{align}
  %
  which are formally of two-loop order.  The method is imprecise if
  the self-energy corrections to the mass matrix are large or the
  tree-level mass spectrum of the multiplet is very split.

@@

\item \code{HighPrecision}: This option solves
  Eq.~\eqref{eq:pole-mass-def} exactly by iterating over the momentum
  $p$.  It therefore provides the determination of the pole masses
  with highest precision, but has also the highest execution time.
  Here the one-loop mass matrix $m_{f,1L}^\text{high}$ is diagonalized
  $n$ times, as in the case of \code{MediumPrecision},
  resulting in $n$ pole masses $m_{f_k}^\pole$ ($k = 1,\ldots,n$).
  Afterwards, the diagonalization is repeated, this time using the
  calculated pole masses $m_{f_k}^\pole$ for the momentum calculation
  $p^2 = (m_{f_k}^\pole)^2$.  The iteration stops if convergence is
  reached.
\end{itemize}
%
A numerical comparison of the three different methods for a specific
CMSSM parameter point can be found in
\tabref{tab:pole-mass-precision-comparison}.  One finds that (i) the
calculated lightest CP-even Higgs pole mass is sensitive to the chosen
pole mass calculation method due to the large loop corrections.
Especially, if two-loop contributions to the Higgs tadpoles and
self-energies are added, the two approximations \code{LowPrecision}
and \code{MediumPrecision} must not be used, because they neglect
terms of two-loop order.  (ii) Since the gluino does not mix with
other particles, there is no difference between \code{LowPrecision}
and \code{MediumPrecision}.  Not neglecting the two-loop terms by
using \code{HighPrecision} changes the gluino mass about $1\unit{\%}$.
(iii) Since the up-type squarks form a $6$-plet, the calculation of
their pole masses is dramatically increased from $2.57\unit{ms}$ to
$60.9\unit{ms}$ between \code{LowPrecision} and \code{HighPrecision}.
However, since the change in the lightest up-type squark mass between
the three different methods is only of the order $\unit{0.4\%}$, one
can consider calculating then with \code{MediumPrecision} only.
%
\begin{table}[tbh]
  \centering
  \begin{tabular}{llll}
    \toprule
    & \code{LowPrecision}
    & \code{MediumPrecision}
    & \code{HighPrecision}\\
    \midrule
    $m_h$
    & $1441\unit{GeV}$ ($0.96\unit{ms}$)
    & $128\unit{GeV}$ ($2.53\unit{ms}$)
    & $126\unit{GeV}$ ($9.79\unit{ms}$) \\
    $m_{\tilde{g}}$
    & $10167\unit{GeV}$ ($0.10\unit{ms}$)
    & $10167\unit{GeV}$ ($0.10\unit{ms}$)
    & $10251\unit{GeV}$ ($0.41\unit{ms}$) \\
    $m_{\tilde{u}_1}$
    & $8702\unit{GeV}$ ($2.57\unit{ms}$)
    & $8714\unit{GeV}$ ($15.7\unit{ms}$)
    & $8734\unit{GeV}$ ($60.9\unit{ms}$) \\
    \bottomrule
  \end{tabular}
  \caption{Comparison of pole masses with different calculation
    methods for the CMSSM with $m_0=8\unit{TeV}$, $M_{1/2}=A_0=5\unit{TeV}$,
    $\tan\beta=20$, $\sign\mu=+1$.  In brackets the time used to calculate
    the pole masses of the whole multiplet is given.}
  \label{tab:pole-mass-precision-comparison}
\end{table}
%

>>>>

The (physical) pole mass spectrum can now be calculated.  \fs uses the
full one-loop self-energies and tree-level mass matrices obtained from
SARAH to calculate the pole masses, which means finding the values $p$
that solve the equation
%
\begin{align}
  0 = \det\left[p^2\unity - m_{f,1L}(p^2)\right].
  \label{eq:pole-mass-def}
\end{align}
%
Here the one-loop mass matrix $m_{f,1L}(p^2)$ for field $f$ is given
in terms of the tree-level mass matrix $m_f$ and the self-energy
$\Sigma_f(p^2)$ as
%
\begin{align}
  &\text{scalars } \phi: &
  m_{\phi,1L}(p^2) &= m_{\phi} - \Sigma_\phi(p^2), \\
  &\text{Majorana fermions } \chi: &
  m_{\chi,1L}(p^2) &= m_{\chi} - \frac{1}{2}\Big[
    \Sigma_\chi^S(p^2) + \Sigma_\chi^{S,T}(p^2)
    + \Big( \Sigma_\chi^{L,T}(p^2) + \Sigma_\chi^R(p^2) \Big) m_{\chi} \notag \\
    &&&\phantom{= m_{\chi} - \frac{1}{2}\Big[}
    + m_{\chi} \Big( \Sigma_\chi^L(p^2) + \Sigma_\chi^{R,T}(p^2) \Big)
  \Big], \label{eq:majorana-loop-corrections} \\
  &\text{Dirac fermions } \psi: &
  m_{\psi,1L}(p^2) &= m_{\psi}
  - \Sigma_\psi^S(p^2)
  - \Sigma_\psi^R(p^2) m_{\psi}
  - m_{\psi} \Sigma_\psi^L(p^2) .
\end{align}
%
Eq.~\eqref{eq:pole-mass-def} can be solved by diagonalizing the
one-loop mass matrix $m_{f,1L}(p^2)$.  However, since $m_{f,1L}(p^2)$
depends on the momentum $p$, an iteration over $p$ must be performed.
Since this iteration can be very time consuming for large field
multiplets, \fs provides two approximative procedures with a shorter
run-time in addition to the iterative procedure.  In the \fs model
file the two approximative procedures are called \code{LowPrecision}
and \code{MediumPrecision}.  The iterative procedure is called
\code{HighPrecision}.  The procedure to be used can be set in the
model file for each field.  The default setting is
%
\begin{lstlisting}
DefaultPoleMassPrecision = MediumPrecision;
HighPoleMassPrecision    = {hh, Ah, Hpm};
MediumPoleMassPrecision  = {};
LowPoleMassPrecision     = {};
\end{lstlisting}
%
In the variable \code{DefaultPoleMassPrecision} the default pole mass
calculation precision to be used for all particles is selected.
Possible values are \code{LowPrecision}, \code{MediumPrecision} and
\code{HighPrecision}.  The values \code{LowPrecision} and
\code{MediumPrecision} correspond to the two approximations described
below and \code{HighPrecision} corresponds to the iterative
determination of the pole masses.  In the variables
\code{HighPoleMassPrecision}, \code{MediumPoleMassPrecision} and
\code{LowPoleMassPrecision} the pole mass calculation precision can be
changed for individual particles.  The settings in these variables
overwrite the setting in \code{DefaultPoleMassPrecision} for these
particles.  In the above example the pole masses of all particles are
calculated with \code{MediumPrecision}, except for the Higgs boson
pole masses, which are calculated with the iterative procedure
(\code{HighPrecision}).

The three different pole mass calculation procedures work as follows:
%
\begin{itemize}
\item \code{LowPrecision}: This option provides the lowest precision
  but is also the fastest one.  Here the one-loop mass matrix
  $m_{f,1L}^\text{low}$ is calculated exactly once as
%
  \begin{align}
    \forall i,j: (m_{f,1L}^\text{low})_{ij} = (m_{f,1L}(p^2 = m_{f_i}
    m_{f_j}))_{ij} ,
  \end{align}
%
  where $m_{f_i}$ is the $i$th mass eigenvalue of the tree-level mass
  matrix $m_f$.  Afterwards, $m_{f,1L}^\text{low}$ is diagonalized and
  the eigenvalues are interpreted as pole masses $m_{f_i}^\pole$.
  This method neglects terms of the form
  %
  \begin{align}
    \left[(m_{f_k}^\pole)^2 - m_{f_i}m_{f_j}\right]
    \left.\frac{\partial m_{f,1L}(p^2)}{\partial
        p^2}\right|_{p^2=m_{f_i}m_{f_j}} ,
  \end{align}
  %
  which are formally of two-loop order.  The method is imprecise if
  the self-energy corrections to the mass matrix are large or the
  tree-level mass spectrum of the multiplet is very split.  Note: We
  strongly discourage the use of this method for the determination the
  Higgs pole masses, as the result will very imprecise due to the
  large loop corrections.  \fs will print a warning if this method is
  used for any Higgs boson.

@@

\item \code{HighPrecision}: This option solves
  Eq.~\eqref{eq:pole-mass-def} exactly by iterating over the momentum
  $p$.  It therefore provides the determination of the pole masses
  with highest precision, but has also the highest execution time.
  Here the one-loop mass matrix $m_{f,1L}^\text{high}$ is diagonalized
  $n$ times, as in the case of \code{MediumPrecision},
  resulting in $n$ pole masses $m_{f_k}^\pole$ ($k = 1,\ldots,n$).
  Afterwards, the diagonalization is repeated, this time using the
  calculated pole masses $m_{f_k}^\pole$ for the momentum calculation
  $p^2 = (m_{f_k}^\pole)^2$.  The iteration stops if convergence is
  reached.
A numerical comparison of the three different methods for a specific
CMSSM parameter point can be found in
\tabref{tab:pole-mass-precision-comparison}.  One finds that
(i) the calculated lightest CP-even Higgs pole mass, $m_h$, differs
about $2\unit{GeV}$ between \code{MediumPrecision} and
\code{HighPrecision}, due to large loop corrections.  Since the
experimental Higgs mass uncertainty is currently around
$0.4\unit{GeV}$ \cite{Beringer:1900zz}, we strongly recommend the use
of \code{HighPrecision} to calculate the Higgs pole mass (this is the
default).  Especially, if two-loop contributions to the Higgs tadpoles
and self-energies are added \code{MediumPrecision} must not be used,
because it neglects terms of two-loop order.  The Higgs boson mass for
the \code{LowPrecision} method is not given in the table, as it will
lead to an imprecise result and is therefore strongly discouraged to
be used, see \secref{sec:PoleMasses}.
(ii) The gluino pole mass, $m_{\tilde{g}}$, is given in the second
row.  Since the gluino does not mix with other particles, there is no
difference between \code{LowPrecision} and \code{MediumPrecision}.
Not neglecting the two-loop terms by using \code{HighPrecision}
increases the gluino mass about $1\unit{\%}$ for this parameter point.
(iii) The pole mass of the lightest neutralino,
$m_{\tilde{\chi}_1^0}$, is given in the third row.  Since the
momentum-dependent loop-corrections to the lightest neutralino mass
are small for this parameter point, it's pole mass varies only in the
sub-GeV range between the three methods.  However, the run-time of the
\code{LowPrecision} method is more than a factor $10$ smaller than of
the \code{HighPrecision}, due to the complicated structure of the loop
corrections in Eq.~\eqref{eq:majorana-loop-corrections}.
(iv) The pole masses of the lightest sfermions are given in rows
$4$--$6$.  Since these particles are contained in $6$-plets, the
run-time for the calculation of their pole masses is dramatically
increased by around a factor $20$ between \code{LowPrecision} and
\code{HighPrecision}.  However, since the change in the lightest
sfermion masses between the three different methods is less than
$\unit{0.4\%}$, one can consider calculating then with
\code{MediumPrecision} only.
\begin{table}[tbh]
  \begin{tabular}{lrrr}
    \toprule
    & \code{LowPrecision}
    & \code{MediumPrecision}
    & \code{HighPrecision}\\
    \midrule
    $m_h$
    & --
    & $128.3\unit{GeV}$ ($2.45\unit{ms}$)
    & $126.3\unit{GeV}$ ($9.53\unit{ms}$) \\
    $m_{\tilde{g}}$
    & $10167\unit{GeV}$ ($0.10\unit{ms}$)
    & $10167\unit{GeV}$ ($0.10\unit{ms}$)
    & $10251\unit{GeV}$ ($0.38\unit{ms}$) \\
    $m_{\tilde{\chi}_1^0}$
    & $2295\unit{GeV}$ ($4.17\unit{ms}$)
    & $2295\unit{GeV}$ ($16.1\unit{ms}$)
    & $2295\unit{GeV}$ ($64.2\unit{ms}$) \\
    $m_{\tilde{u}_1}$
    & $8703\unit{GeV}$ ($2.45\unit{ms}$)
    & $8714\unit{GeV}$ ($14.9\unit{ms}$)
    & $8734\unit{GeV}$ ($58.5\unit{ms}$) \\
    $m_{\tilde{d}_1}$
    & $10442\unit{GeV}$ ($2.48\unit{ms}$)
    & $10391\unit{GeV}$ ($15.0\unit{ms}$)
    & $10418\unit{GeV}$ ($59.9\unit{ms}$) \\
    $m_{\tilde{e}_1}$
    & $7913\unit{GeV}$ ($1.94\unit{ms}$)
    & $7909\unit{GeV}$ ($11.2\unit{ms}$)
    & $7909\unit{GeV}$ ($34.1\unit{ms}$) \\
    \bottomrule
  \end{tabular}
  \caption{Comparison of pole masses with different calculation
    methods for the CMSSM with $m_0=8\unit{TeV}$, $M_{1/2}=A_0=5\unit{TeV}$,
    $\tan\beta=20$, $\sign\mu=+1$.  In brackets the time used to calculate
    the pole masses of the whole multiplet is given.}
  \label{tab:pole-mass-precision-comparison}
\end{table}
